---
title: "Vignette Title"
author: "Brian O'Meara"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rphylotastic usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`rphylotastic` connects to Phylotastic web services, which themselves wrap and process multiple other services as well as have unique services of their own.

# Getting trees

The main use for `rphylotastic` is getting phylogenetic trees. These can come from Open Tree of Life (topology only, many polytomies), Phylomatic (has branch lengths, but only plants), or NCBI taxonomy (even more polytomies than Open Tree's tree). See also the `treebase` and `rotl` packages for getting individual trees, or `datelife` for getting chronograms (a single synthetic one or individual ones).

```{r}
taxa <- c("Crabronidae", "Ophiocordyceps", "Megalyridae",
          "Formica polyctena", "Tetramorium caespitum",
          "Pseudomyrmex", "Carebara diversa", "Formicinae")
phy <- taxa_get_otol_tree(taxa)
plot(phy)
```



# Citing

The phylotastic manuscript is still being written.
