---
title: "rphylotastic package"
author: "Brian O'Meara and Luna L. Sánchez-Reyes"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rphylotastic usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, warnings = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "# >"
)
```

`rphylotastic` connects to Phylotastic web services, which themselves wrap and process multiple other services as well as have unique services of their own.

You first need to install and load the package:

CRAN version:

```{r cran, include=TRUE, eval=FALSE}
install.packages("rphylotastic")
```

Or development version:

```{r devtools, include=TRUE, eval=FALSE}
library(devtools)
devtools::install_github("phylotastic/rphylotastic")
```

Then load the package:

```{r load, include=FALSE}
library(rphylotastic)
```

# Getting trees

The main use for `rphylotastic` is getting phylogenetic trees for a set of taxa of interest. Trees can come from Open Tree of Life (topology only, many polytomies), Phylomatic (has branch lengths, but only plants and mammals), or NCBI taxonomy (even more polytomies than Open Tree's tree). See also the `treebase` and `rotl` packages for getting individual trees, or `datelife` for getting chronograms (a single synthetic one or individual ones).

If we would like to know about north american endemic carnivorous plants such as the venus flytrap (_Dionaea muscipula_), pitcher plants (_Sarracenia_ genus) and the cobra lily (_Darlingtonia californica_) in relation to other plants with similar behaviour such as the popular _Drosera_ sundues, _Pinguicula_ butterworts and the aquatic carnivorous _Utricularia_ bladderworts, which have a wider range of distribution:

```{r, fig.align = 'center', fig.retina = 2, fig.margin = TRUE, fig.width = 6}

taxon_names <- c("Dionaea muscipula", "Sarracenia", "Darlingtonia californica",
          "Drosera", "Pinguicula", "Utricularia", "Roridulaceae")
phy1 <- taxa_get_otol_tree(taxa = taxon_names)

phy2 <- taxa_get_phylomatic_tree(taxa = taxon_names)

# phy3 <- taxa_get_taxonomic_tree(taxa = taxon_names) # this service is giving Error in open.connection(con, "rb") : HTTP error 500. (2018.06.21)

source_tree <- c("Open Tree of Life", "Phylomatic")
# source_tree <- c("Open Tree of Life", "Phylomatic", "NCBI taxonomy")

trees <- c(phy1, phy2)
# trees <- c(phy1, phy2, phy3)

library(ape)
par(xpd = FALSE, mfrow = c(1, length(trees)), mai = c(0.1, 0.1, 0.5, 0.1), oma = c(0, 1, 2, 1))
for (i in 1:length(trees)){
  plot(ladderize(trees[[i]]), cex = 0.5, label.offset = 0.3)
  mtext(paste("from", source_tree[i]), cex = 1, side = 3, line = 1)
}
mtext("Carnivorous plant tree", outer = T, side = 3, cex = 1.5, line =0)

```

With `rphylotastic`, you can also check if any two trees are exactly equal.
```{r}
phylo_compare(phy1, phy1)
phylo_compare(phy1, phy2)
```

# Getting scientific names
## From scratch
One very cool feature of `rphylotastic` is the ability to scrape scientific names from documents or web pages. For example, you can get all latin names from a wikipedia article on primates:

```{r}
sci_names <- url_get_scientific_names(URL = "https://en.wikipedia.org/wiki/Primate")
length(sci_names)
head(sci_names)
```

You can also extract all scientific names from a text you are studying:
```{r}
sci_names <- text_get_scientific_names(text = "The domestic cat (Felis silvestris catus or Felis catus) is a small, typically furry, carnivorous mammal.")
head(sci_names)
```


## From a taxon of interest
If you are interested in a particular group, and you want to get a tree containing all species within the group, function `taxon_get_species` will get you a vector of all species registered in Open Tree of Life taxonomy for that group. You can then use that vector to perform a tree search.

```{r}
ant_species <- taxon_get_species(taxon = "Formicidae")
more_species <- taxon_get_species(taxon = c("Formicidae", "Felidae"))
length(more_species)
length(ant_species)
head(ant_species)
# phy1 <- taxa_get_otol_tree(taxa = ant_species)
```
You can only use this function one taxon at a time
```{r}
more_species <- taxon_get_species(taxon = c("Formicidae", "Felidae"))
length(more_species)
```
But you can use a handy apply function
```{r}
more_species <- rapply(list("Formicidae", "Felidae"), taxon_get_species)
length(more_species)
```
Or see `datelife` package for a wrapper of this function accepting multiple taxon names at a time.

### Subsetting species names
Species from a taxon can also be subset by country or by genome:
```{r}
spp_from_country <- taxon_get_species_from_country("Formicidae", "Mexico")
length(spp_from_country)
head(spp_from_country)

spp_with_genome <- taxon_get_species_with_genome("Formicidae")
length(spp_with_genome)
head(spp_with_genome)
```
# Checking your scientific names
You may already have a list of names, but you want to check if they all are good scientific names. `rphylotastic` calls Global Name Resolver and the Taxonomic Name Resolution Service to verify names. This function will keep correct names only.

```{r}
species <- c("Delphinus delphus", "Homo sepiens")
taxa_resolve_names_with_otol(taxa = species)
```
Alternatively, you can use services from `rotl` and `taxize` packages to clean your names:

```{r}
library(taxize)
species_gnr <- gnr_resolve(names = species)
nrow(species_gnr)
# it is a big table. Let's subset by unique matched names, so it is easier to choose:
species_gnr[!duplicated(species_gnr$matched_name),c("user_supplied_name", "matched_name")]

tnrs(query = species)
```
On the example above, species names were matched to iPlant database only. If you look at the score, it's pretty low (<0.6), so you can tell that this match is probably wrong.
```{r}
library(rotl)
species_tnrs <- tnrs_match_names(names = species)
species_tnrs[,1:5]
```
Package `brranching` has a very nice function that allows choosing among multiple matches. This function is actually used in phylomatic tree searches implemented via `rphylotastic` function.

```{r, eval = FALSE}
library(brranching)
phylomatic_names("Byblis")
# [1] "byblidaceae/byblis/byblis"
```

# Getting data from Encyclopedia of Life
You can also get some images:

```{r}
some_images <- species_get_image_data(species = c("Homo sapiens", "Felis catus"))
```
Or other miscelanous info:
```{r}
some_info <- species_get_info(species = c("Homo sapiens", "Felis catus"))
```

# Managing your lists of names
You can get an phylotastic! user ID by registering with your gmail address on the portal. Then, you can get a temporary token sent to you email to help you manage your lists from the R console. Here are a few examples, but they won't work for you since th etoken expire afer a few hours.
```{r, eval = FALSE}
get_species_from_list()
update_species_in_list()
insert_species_in_list()
replace_species_in_list()
remove_species_from_list()
```

# Citing

The phylotastic manuscript is still being written.
